### 2023-10-31 journal/correctness-GC.tex
editing Sec 6

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then it has been propagated to the root and its result has been written in its \fld{response} field.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.

@@

\here{need to prove a lemma analogous to Lemma \ref{blockSize}}

@@

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $2q_{max}+4p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

@@

The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
Any enqueue in $L_2$ whose value is not still in the queue at the end of $L_2$
must still be in progress at~$C$; otherwise the process that dequeued it would have
set its \var{last} entry to a value greater than the index of the block of the root that contains the enqueue prior to $C$,
contradicting the definition of $m$.
So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ enqueues in $L_2$.

If a dequeue in $L_2$ returns a non-null value in the sequential execution $L$,
then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
Thus, there are at most $q_{max}+(q_{max}+p)$ dequeues in $L_2$ that return non-null values.
Any dequeue in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
at $C$; otherwise the process that performed the dequeue would have set its \var{last} entry
to a value greater than the index of the block of the root that contains the dequeue prior to $C$, 
contradicting the definition of $m$.
So, there are at most $p$ null dequeues in $L_2$.
Thus, there are at most $2q_{max} + 2p$ dequeues in $L_2$.

$A$'s call to \op{SplitBlock} determines the index $s$ used to split \var{v}'s RBT by following 
\eleft\ and \eright\ pointers from the root down to \var{v}.
So, the block returned is a subblock of block $m$ of the root, unless
at some point along the path of subblocks the subblock has already been removed by a split,
in which case \op{SplitBlock} returns a subblock of a block $m'>m$ in the root.

Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$  sets
$T'\assign\op{Split}(T,s)$.
Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$.\here{cite a lemma about propagation before operation terminates}
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
There are at most $p$ operations in block $m$ of the root 
and we showed above that there are at most $2q_{max} + 2p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $2q_{max}+4p$ operations in blocks of $T'$.
Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $2q_{max}+4p$ blocks, and one more 
block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
\end{proof}

@@

The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
\here{For later version:  There may be room to improve this bound by a more careful analysis since the bounds so far
didn't use the fact that operations only appear along one path in the tree; in other words
doing a bound on total size of all RBTs at one level of the ordering tree at once, rather than
on each node separately, might reduce it to qlogp + poly(p).  Might also be possible to be
more precise about how it depends on "current" size of queue rather than max size.}

@@

\here{It may be possible, via a more careful analysis, to reduce the bound on space usage to $O(\log^2 p(q_{max}+p^3)$.}

Although individual operations may now become more expensive because they have to perform GC,
we show that operations still have polylogarithmic amortized step complexity.

\begin{mytheorem}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{mytheorem}
\begin{proof}
\here{this proof may need some modification since code was changed to change how split point is computed}
It follows from Corollary \ref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can  be done in $O(\log(p+q_{max}))$ steps.

@@

Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.
If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th helping phase on \var{v}}.

We argue that each process $P$ can do an $r$th helping phase on \var{v} at most once.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th helping phase on \var{v}.
Let $\var{v},T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
Either way, by Lemma \ref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th helping phase on \var{v} again.

Each helping phase takes $O(p)$ steps to read the \var{last} array, and
$O(p \log p \log(p+q_{max}))$ steps in \op{Help},
and $O(\log(p+q_{max}))$ steps to split and insert a new node into a RBT.
Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps
are performed by all processes during their $r$th helping phase on \var{v}.
We can amortize these steps over the operations that appear in 
$\var{v}.\fld{blocks}[(r-1)G+1..rG]$.
By Corollary \ref{blockNotEmpty}, there are at least $G$ such operations, 
so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$.
\end{proof}

The implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
Since each operation still  does only two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.

>>>>

\begin{lemma}\label{blocknotfound}
If a \op{Dequeue} operation fails to find a block in an RBT,
%during its calls to \op{IndexDequeue} or \op{GetEnqueue}, 
then it has been propagated to the root and its result has been written in the \fld{response} field of the leaf block it created.
If an \op{Enqueue} operation fails to find a block in an RBT, it has been propagated to the root.
\end{lemma}
\begin{proof}
Any block that GC removes from an RBT is finished, by Invariant \ref{inv::finished}, so if an 
\op{Enqueue} or \op{Dequeue} fails to find a block while it is propagating itself up to
the root (for example, during the \op{CreateBlock} routine), then a block containing
the operation itself has been removed from a RBT, so the operation has propagated to the root,
by the definition of finished.
Moreover, by Invariant \ref{inv::finished}, if the operation is a dequeue, then its result is in its \fld{response} field.

@@

Lemma \ref{blockSize} shows that each block contains at most one operation of each process.
Its proof depends only on Lemma \ref{lem::subblocksDistinct}, Corollary \ref{blockNotEmpty} and Lemma \ref{lem::appendExactlyOnce}, which are all still true for the space-bounded implementation.  So the same proof of Lemma \ref{blockSize} still applies.

@@

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $3q_{max}+5p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let $C$ be the configuration before $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $C$.

@@

The number of enqueues in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
If the value enqueued by any enqueue in $L_2$  is not still in the queue
at the end of $L_2$, then the dequeue in $L_2$ that dequeued that value must still be in progress
at $C$; otherwise the process that performed that dequeue would have set its \var{last} entry
to the index of the root block that contains the enqueue, which is greater than $m$, before $C$, contradicting
the fact that all values in the $last$ array at $C$ are less than or equal to $m$.
So, there are at most $p$ enqueues in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ enqueues in $L_2$.

If a dequeue in $L_2$ returns a non-null value in the sequential execution $L$,
then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
Thus, there are at most $q_{max}+(q_{max}+p)$ dequeues in $L_2$ that return non-null values.
Any dequeue in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
at $C$; otherwise the process that performed the dequeue would have set its \var{last} entry
to a value greater than the index of the root block that contains the dequeue prior to $C$, 
contradicting the definition of $m$.
So, there are at most $p$ null dequeues in $L_2$.
Thus, there are at most $2q_{max} + 2p$ dequeues in $L_2$.

$A$'s call to \op{SplitBlock} determines the index $s$ used to split \var{v}'s RBT by following 
\eleft\ and \eright\ pointers from the root down to \var{v}.
So, the block returned is a subblock of the root block with index $m$, unless
at some point along the path of subblocks the subblock has already been removed by a split,
in which case \op{SplitBlock} returns a subblock of a root block with index $m'>m$.

Next, we bound the number of operations in \var{v}'s blocks that are retained when $A$  sets
$T'\assign\op{Split}(T,s)$.
Since $T$ was read before $C$, any operation in $T$ is either in progress at $C$ or has been propagated to the root before $C$, by Lemma \ref{lem::appendExactlyOnce}.
Thus, there are at most $p$ operations in $T$ that do not appear in $L_1\cdot L_2$.
All the rest of the operations in blocks of $T'$ have been propagated to blocks $m..\ell$ of the root.
By Lemma \ref{blockSize}, There are at most $p$ operations in block $m$ of the root 
and we showed above that there are at most $3q_{max} + 3p$ in blocks 
$m+1..\ell$ of the root.
Thus, there are at most $3q_{max}+5p$ operations in blocks of $T'$.
Since each block is non-empty by Corollary \ref{blockNotEmpty}, $T'$ contains at most $3q_{max}+5p$ blocks, and one more 
block is inserted before $A$ sets \var{v}'s \fld{blocks} to the resulting RBT.
\end{proof}

@@

The following theorem bounds the space that is reachable (and therefore cannot be freed by the environment's garbage collector) at any time.
%\here{At first, I thought the following analysis could be tightened
%by using the fact that operations only appear along one path in tree
%so doing a bound on total size of all nodes on one level might work yield better bound.
%But I don't think so, because we could have each process go in order to dequeue
%everything and then enqueue q new items (where q>>p).  Then sets of enqueue operations
%stored in each node at one level of tree would be disjoint (because of the way
%GC is timed differently at each node).

@@

Performing GC on a node takes $\Theta(p \log p\log(q_{max}+p))$ steps in the worst case (as explained in the following proof),
so an individual operation can take up to $\Theta(p \log^2 p\log(q_{max}+p))$ steps
if it helps at each node along the path from a leaf to the root.
However,
we show that operations still have polylogarithmic amortized step complexity.

\begin{mytheorem}
The amortized step complexity of each operation is $O(\log p \log(p+q_{max}))$.
\end{mytheorem}
\begin{proof}
It follows from Corollary \ref{RBTbound} and our choice of $G=p^2\ceil{\log p}$
that all the RBT routines we use to perform \op{Split}, \op{Insert} and searches for
blocks with a particular index or for a \fld{sum\sub{enq}} value (in line \ref{FRsearchGC} or \ref{getChildGC}) can  be done in $O(\log(p+q_{max}))$ steps.

@@

Now we consider the additional steps a process takes while doing GC in line \ref{GCstart}--\ref{GCend}
and show that the amortized number of GC steps each operation performs is also $O(\log p\log(p+q_{max}))$.
If a process does GC in a call to \op{AddBlock}($\var{v},T,B$) where $B$ has \fld{index} $r\cdot G$ for some integer $r$, we call this the process's \emph{$r$th GC phase on \var{v}}.

We argue that each process $P$ can do an $r$th GC phase on \var{v} at most once.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th GC phase on \var{v}.
Let $\var{v},T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
Either way, by Lemma \ref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th GC phase on \var{v} again.

Each GC phase takes $O(p)$ steps for \op{SplitBlock} to read the \var{last} array and figure
out where to split the $blocks$ RBT, 
$O(p \log p \log(p+q_{max}))$ steps in \op{Help},
and $O(\log(p+q_{max}))$ steps to split and insert a new node into the RBT.
Thus, for each integer $r$ and each node \var{v}, a total of $O(p^2\log p\log(p+q_{max}))$ steps
are performed by all processes during their $r$th helping phase on \var{v}.
We can amortize these steps over the operations that appear in 
$\var{v}.\fld{blocks}[(r-1)G+1..rG]$.
By Corollary \ref{blockNotEmpty}, there are at least $G$ such operations, 
so each operation's amortized number of steps for GC at each node along the path from its leaf to the root
is $O(p^2\log p\log(p+q_{max})/G)=O(\log(p+q_{max}))$.
Hence each operation's amortized number of GC steps is $O(\log p\log(p+q_{max}))$.
\end{proof}

The implementation remains wait-free:  the depth of recursion in each routine is  bounded
by the height of the tree and the only loop is the counted loop in the \op{Help} routine.
Moreover, since each operation still  does only two \op{CAS} instructions at each level of the tree (at line \ref{casGC}), the following proposition still holds for the space-bounded version of the queue.
